Semantic integration of physiology phenotypes with an application to the Cellular Phenotype Ontology

### 2012-03-25 paper/drinking-bioinformatics.tex

\section{Results:}
Here, we present the outline of a theory and its implementation for an
ontology of physiology-related phenotypes. We provide a formal
description of process attributes and relate them to the attributes of
their temporal parts and participants. We apply our theory to create
the Cellular Phenotype Ontology (CPO). The CPO is an ontology of
morphological and physiological phenotypic characteristics of cells,
cell components and cellular processes. Its prime application is the
unification of cellular phenotype descriptions across species by
providing terms and uniform definition patterns. The CPO can further
be used for the annotation of observed abnormalities in domains, such
as systems microscopy, in which cellular abnormalities are observed
and for which no phenotype ontology has been created.

@@

PATO allows for the description of affected entities by combining
various ontologies that describe the entities that have been affected,
such as the various anatomical ontologies, GO \cite {Ashburner2000short},
the Cell Type Ontology \cite {Bard2005}, SO \cite {Eilbeck2005} etc
with the various qualities it provides for defining how these entities
were affected.  PATO can be used for annotation either directly in a
so called post-composed (post-coordinated) manner or for providing
formal (logical) definitions (equivalence axioms) to ontologies
containing a set of precomposed (pre-coordinated) phenotype terms. For
instance, to describe the decrease in the length of the sexual cycle
of female animals, we can combine the PATO term \textit{decreased
  duration} ({\tt PATO:0000499}) with the Gene Ontology term
\textit{estrous cycle} ({\tt GO:0042698}), whilst if such a term
existed in a pre-composed ontology (for example the {\tt MP:0009007}
term from the Mammalian Phenotype) it could be used to provide an
equivalence statement between that class and the above PATO-based
description.

@@

Our three aims rely on the possibility of using automated reasoning
over a resulting ontology of physiology phenotypes.  However, OWL is
an expressive formal language, and automated reasoning in OWL has a
high computational complexity (reasoning in OWL 2.0 is
2NEXPTIME-complete). Consequently, due to the exponential increase in
worst-case time complexity for automated reasoning in OWL, we would
not be guaranteed to achieve our aims once we consider more than very
few phenotype classes. In particular, using an ontology of the size of
GO, with more than XX,XXX classes for processes, as a foundation for
constructing an ontology of physiology phenotypes would not allow us
to achieve our aims through automated reasoning.

>>>>

\section{Results:}
Here, we present the outline of a theory and its implementation for an
ontology of physiology-related phenotypes. We provide a formal
description of process attributes and relate them to the attributes of
their temporal parts and participants. We apply our theory to create
the Cellular Phenotype Ontology (CPO). The CPO is an ontology of
morphological and physiological phenotypic characteristics of cells,
cell components and cellular processes. Its prime application is to
provide terms and uniform definition patterns for the annotation of
cellular phenotypes. The CPO can be used for the annotation of
observed abnormalities in domains, such as systems microscopy, in
which cellular abnormalities are observed and for which no phenotype
ontology has been created.

@@

PATO allows for the description of affected entities by combining
various ontologies that describe the entities that have been affected,
such as the various anatomical ontologies, GO \cite {Ashburner2000short},
the Cell Type Ontology \cite {Bard2005}, SO \cite {Eilbeck2005} etc
with the various qualities it provides for defining how these entities
were affected.  PATO can be used for annotation either directly in a
so called post-composed (post-coordinated) manner or for providing
formal (logical) definitions (equivalence axioms) to ontologies
containing a set of precomposed (pre-coordinated) phenotype terms. For
instance, to describe the decrease in the length of the sexual cycle
of female animals, we can combine the PATO term \textit{Decreased
  duration} ({\tt PATO:0000499}) with the Gene Ontology term
\textit{Estrous cycle} ({\tt GO:0042698}), whilst if such a term
existed in a pre-composed ontology (for example the {\tt MP:0009007}
term from the Mammalian Phenotype) it could be used to provide an
equivalence statement between that class and the above PATO-based
description.

@@

Our three aims rely on the possibility of using automated reasoning
over a resulting ontology of physiology phenotypes.  However, OWL is
an expressive formal language, and automated reasoning in OWL has a
high computational complexity (reasoning in OWL 2.0 is
2NEXPTIME-complete). Consequently, due to the exponential increase in
worst-case time complexity for automated reasoning in OWL, we would
not be guaranteed to achieve our aims once we consider more than very
few phenotype classes. In particular, using an ontology of the size of
GO, with more than 22,000 classes for processes, as a foundation for
constructing an ontology of physiology phenotypes would not allow us
to achieve our aims through automated reasoning.
